Modelling circumstellar discs with 3D radiation hydrodynamics

### 2009-08-27 SPH-Torus-paper1/sphtorus_benchmark.tex
Minor changes.
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2026 25464f21-6099-402c-b30a-640f1d40d38b

We present results from combining a grid-based radiative transfer code
with a Smoothed Particle Hydrodynamics code to produce a flexible
system for modelling radiation hydrodynamics. The system is tested
using a benchmark model of a circumstellar disc. We find that the SED
and temperature distribution within the disc are highly sensitive to
the representation of the disc inner edge, which depends critically on
both the grid and SPH resolution. The system is then used to model a
circumstellar disc around a T-Tauri star and is found to produce a
steady state disc in both hydrostatic and radiative equilibrium.

@@

This paper presents a results from coupling the {\small{TORUS}} radiative
transfer code \citep{harries_2000} with the SPH (Smoothed Particle
Hydrodynamics) code of \cite{bate_2009}.  {\small{TORUS}} performs radiative
transfer calculations using the Monte-Carlo method of \cite{lucy_1999}
with a flux limited diffusion approximation in regions of high optical
depth. The radiative transfer calculations are performed on a grid
which uses adaptive mesh refinement (AMR) to allow variable
resolution.  The SPH method is well suited to dealing with the large
range in size scales encountered during star formation processes, hence
an SPH code coupled to a combined Monte-Carlo and flux limited
diffusion radiative transfer code is a very flexible method for
performing star formation calculations incorporating radiative
feedback.

@@

Accurately coupling the SPH component to the radiative transfer
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
benchmark case (see Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The spectral energy distribution
and temperature distribution of the benchmark disc are examined in
Section~\ref{section:seds} in order to understand how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc} with the aim of calculating an
equilibrium state in three dimensional hydrostatic and radiative
balance.

@@

An important step in setting up grid-based radiative transfer using a
density distribution derived from SPH particles is determining an
effective transformation from the Lagrangian particle description to
the Eulerian grid desciption.  The representation on grid should
preserve important properties of the SPH model (e.g. total mass) but
also needs to take account of the possibility that a radiative
transfer calculation requires the highest resolution in different
regions to a hydrodynamics calculation (e.g. to resolve opacity
gradients). This section determines a grid construction method which
allows a good degree of control over the number of cells in the grid,
and the location of the highest resolution regions, while maintaining
an accurate representation of the disc mass.

@@

For some geometries it may be appropriate to use a cylindrical polar
co-ordinate system. A mass per cell splitting condition can still be
used but it needs to be modified to take into account the increase in 
cell volume as cylindrical polar radius increases, for fixed values of
$dr$ and $dz$.  
%
Consider a cartesian grid with cubic volume elements and let
\begin{equation}
dl_{\rm{cart}} = dx = dy = dz 
\end{equation}
so that
\begin{equation}
dl_{\rm{cart}} = \left( \frac{dM_{\rm{cart}}}{\rho_{\rm{cell}}} \right)
^{1/3}
\label{eqn:cart_cell}
\end{equation}
where $dM_{\rm{cart}}$ is the mass of a cell in a cartesian grid and
$\rho_{\rm{cell}}$ is the density of the cell. 
%
In a cylindrical polar grid the mass of a cell is given by
\begin{equation}
dM_{\rm{cyl}} = \rho_{\rm{cell}} \frac{2 \pi r dr dz}{n_{\rm{azi}}}
\end{equation}
if there are $n_{\rm{azi}}$ uniform azimuthal cells. If the 
resolution in $r$ and $z$ is the same then we can define 
\begin{equation}
dl_{\rm{cyl}} = dr = dz 
\end{equation}
which is related to the cell mass and density by
\begin{equation}
dl_{\rm{cyl}} = \left( \frac{n_{\rm{azi}} dM_{\rm{cyl}}}{2 \pi r
  \rho_{\rm{cell}}} \right)  ^{1/2}
\label{eqn:cyl_cell}
\end{equation}
If we require the resolution in cylindrical polar $r$ and $z$
to match the resolution in cartesian $x$, $y$ and $z$ then $dl_{\rm{cyl}} =
dl_{\rm{cart}}$. From eqn~\ref{eqn:cart_cell} and \ref{eqn:cyl_cell} 
\begin{equation}
dM_{\rm{cyl}} = \frac{2 \pi r}{n_{\rm{azi}}} \rho^{1/3} dM_{\rm{cart}}^{2/3}
\end{equation}
which describes the required conversion from a mass per cell limit
suitable for a cartesian grid ($dM_{\rm{cart}}$) to an equivalent mass
per cell limit suitable for a cylindrical polar grid ($dM_{\rm{cyl}}$).

@@

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. {\small{TORUS}} calculates the temperature distribution
using a high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e8}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are AU.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced when $10^6$ particles are used
but is only reduced to a low level with $10^8$ particles.

@@

SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees. At the shortest wavelengths the 12.5 degrees
inclination angle SED is dominated by emission from the central
star. This region is well represented with $10^5$ and $10^8$ particles
but is less well represented with $10^6$ or $10^7$ particles where the
central star is obscured. At the longest wavelengths the disc is optically
thin and the SEDs will be well modelled provided the total disc mass
is accurately represented. In all four cases the total disc mass is
well represented at the long wavelength SED matches the benchmark
results. At intermediate wavelengths the SED is sensitive to the
details of the disc representation. Accurately modelling the SED from
this disc is challenging even with $10^8$ particles. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}

@@

Plots of the disc internal energy are shown in Fig.~\ref{fig:log_u} at siz
different times during the simulation. Figure~\ref{fig:frame1} shows
the initial state before the radiative transfer is switched on. The
disc is vertically isothermal. The second frame (Fig.~\ref{fig:frame2}
is shortly after the raditive transfer has been swithced on. The inner
disc has collapsed and there is heating seen in parts of the disc
exposed to radiation from the central star. In Fig.~\ref{fig:frame3} a
scale height enhancement is present in the disc at a radius of
approximately $3\times10^{15}~\rm{cm}$ which shadows the region
immediately behind causing it to cool. Figure~\ref{fig:frame4} shows
several such scale height enhancements an in each case there is
cooling behind the enhancement. The scale height fluctations are able
to affect the mid-plane temperature due to vertical transport of
radiation. In Fig.~\ref{fig:frame5} the inner region of the disc has
settled and the scale height enhancements are seen only in the outer
disc and by the last frame (Fig.~\ref{fig:frame6}) the disc has
reached a steady state in hydrostatic and radiative equilibrium. 
\begin{figure*}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6675}\label{fig:frame1}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6691}\label{fig:frame2}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6763}\label{fig:frame3}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC7363}\label{fig:frame4}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC9363}\label{fig:frame5}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_run2_DSC1024}\label{fig:frame6}}
  \caption{Particle plots showing internal energy in cyclindrical
    polar co-ordinates. Axis units are cm.}
  \label{fig:log_u}
\end{figure*}

The surface density of the disc is shown in Fig.~\ref{fig:sigma} where
the solid line is the isothermal disc and the dotted line is the disc
at the final time step.  Away from the inner and outer edges of the
disc the surface density is well represented by a power law profile as
a function of radius.
\begin{figure}
  \includegraphics[scale=0.3]{figures/sigma}
  \caption{Surface density as a function of radius.}
  \label{fig:sigma}
\end{figure}
The scale height of the disc was determined by fitting Gaussians to
the vertical density profile in 100 radial bins. The fitted scale
height, as a function of radius, is plotted in
Fig.~\ref{fig:scale_height}.  Within the central 10~AU the density
profile appears flat so it was not possible to fit a meaningful scale
height. The solid line is the isothermal disc and the dotted line is
the disc at the final time step.
\begin{figure}
  \includegraphics[scale=0.3]{figures/scale_height}
  \caption{Disc scale height as a function of radius.}
  \label{fig:scale_height}
\end{figure}
The average temperature within one scale height of the disc mid plane
is plotted in Fig.~\ref{fig:mid_tem}. The solid line is the isothermal
disc and the dotted line is the disc at the final time step.
\begin{figure}
  \includegraphics[scale=0.3]{figures/mid_tem}
  \caption{Average temperature within one scale height of the mid plane as a function of
    radius.}
  \label{fig:mid_tem}
\end{figure}

@@

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a disproportionately large
effect on the temperature distribution in the disc and on the emergent
SED. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles.

The combined SPH-radiative transfer code was able to evolve a
circumstellar disc to a state of radiative and hydrostatic
equilibrium. Fluctuations in scale height were seen, which have also
been observed in codes which combine radiative transfer and vertical
hydrostatic equilibrium calculations. However there were also
transient effects seen in the surface density distribution which would
not be present if the hydrostatic equilibrium was calculated in the
vertical direction only.

By adding self-gravity to the SPH model this method could be used to
explore the effects of radiative transfer on disc
fragmentation. However the model would need to be constructed
carefully to ensure sufficient resolution at the disc inner edge so
that the temperature distribution in the disc is accurate. Such a
model is expected to be feasible although computationally demanding.

>>>>

  We present results from combining a grid-based radiative transfer
  code with a Smoothed Particle Hydrodynamics code to produce a
  flexible system for modelling radiation hydrodynamics. We use a
  benchmark model of a circumstellar disc to determine a robust method
  for constructing a gridded density density distribution from SPH
  particles. The benchmark disc is then used to determine the accuracy
  of the radiative transfer results. We find that the SED and 
  the temperature distribution within the disc are sensitive
  to the representation of the disc inner edge, which depends
  critically on both the grid and SPH resolution. The code is then
  used to model a circumstellar disc around a T-Tauri star. As the
  disc adjusts towards equilibrium vertical motions in the disc are
  induced resulting in scale height enhancements which intecept
  radiation from the central star. Vertical transport of radiation
  enables these perturbations to influence the mid-plane temperature
  of the disc. The vertical motions decay over time and the disc
  ultimately reaches a state of simulatanous hydrostatic and radiative
  equilibrium.

@@

This paper presents a results from coupling the {\small{TORUS}}
radiative transfer code \citep{harries_2000} with the SPH (Smoothed
Particle Hydrodynamics) code of \cite{bate_2009}.  {\small{TORUS}}
performs radiative transfer calculations using the Monte-Carlo method
of \cite{lucy_1999} with a diffusion approximation in regions of high
optical depth. The radiative transfer calculations are performed on a
grid which uses adaptive mesh refinement (AMR) to allow variable
resolution.  The SPH method is well suited to dealing with the large
range in size scales encountered during star formation processes,
hence an SPH code coupled to a Monte-Carlo radiative transfer code is
a very flexible method for performing star formation calculations
incorporating radiative feedback.

@@

Accurately coupling the SPH component to the radiative transfer
component requires an effective transformation from the Lagrangian
particle description of SPH to the Eulerian grid description of an AMR
grid. The procedure used for this transformation is tested using a
cicumstellar disc benchmark (see
Sections~\ref{section:description_of_benchmark}
and~\ref{section:grid_generation}). The spectral energy distribution
and temperature distribution of the benchmark disc are examined in
Section~\ref{section:seds} in order to understand how well the
radiative transfer calculation works with the gridded density field
and to understand the effects of grid and SPH spatial resolution. A
simulation of a circumstellar disc around a T Tauri star is presented
in Section~\ref{section:sph_disc}.

@@

An important step in setting up grid-based radiative transfer using a
density distribution derived from SPH particles is determining an
effective transformation from the Lagrangian particle description to
the Eulerian grid desciption.  The representation on grid should
preserve important properties of the SPH model (e.g. total mass) but
also needs to take account of the possibility that a radiative
transfer calculation requires the highest resolution in different
regions to a hydrodynamics calculation (e.g. to resolve opacity
gradients). This section determines a grid construction method which
allows a good degree of control over the number of cells in the grid,
and the location of the highest resolution regions, while maintaining
an accurate representation of the total disc mass.

@@

For some geometries it may be appropriate to use a cylindrical polar
co-ordinate system. A mass per cell splitting condition can still be
used but it needs to be modified to take into account the increase in 
cell volume as cylindrical polar radius increases, for fixed values of
$dr$ and $dz$.  
%
Consider a cartesian grid with cubic volume elements and let
\begin{equation}
dl_{\rm{cart}} = dx = dy = dz 
\end{equation}
so that
\begin{equation}
dl_{\rm{cart}} = \left( \frac{dM_{\rm{cart}}}{\rho_{\rm{cell}}} \right)
^{1/3}
\label{eqn:cart_cell}
\end{equation}
where $dM_{\rm{cart}}$ is the mass of a cell in the cartesian grid and
$\rho_{\rm{cell}}$ is the density of the cell. 
%
In a cylindrical polar grid the mass of a cell is given by
\begin{equation}
dM_{\rm{cyl}} = \rho_{\rm{cell}} \frac{2 \pi r dr dz}{n_{\rm{azi}}}
\end{equation}
if there are $n_{\rm{azi}}$ uniform azimuthal cells. If the 
resolution in $r$ and $z$ is the same then we can define 
\begin{equation}
dl_{\rm{cyl}} = dr = dz 
\end{equation}
which is related to the cell mass and density by
\begin{equation}
dl_{\rm{cyl}} = \left( \frac{n_{\rm{azi}} dM_{\rm{cyl}}}{2 \pi r
  \rho_{\rm{cell}}} \right)  ^{1/2}
\label{eqn:cyl_cell}
\end{equation}
If we require the resolution in cylindrical polar $r$ and $z$
to match the resolution in cartesian $x$, $y$ and $z$ then $dl_{\rm{cyl}} =
dl_{\rm{cart}}$. From eqn~\ref{eqn:cart_cell} and \ref{eqn:cyl_cell} 
\begin{equation}
dM_{\rm{cyl}} = \frac{2 \pi r}{n_{\rm{azi}}} \rho^{1/3} dM_{\rm{cart}}^{2/3}
\end{equation}
which describes the required conversion from a mass per cell limit
suitable for a cartesian grid ($dM_{\rm{cart}}$) to an equivalent mass
per cell limit suitable for a cylindrical polar grid ($dM_{\rm{cyl}}$).

@@

Figure~\ref{fig:n_part_tem_diff} shows fractional errors in the
temperature distribution for different numbers of
particles. {\small{TORUS}} calculates the temperature distribution
using a high resolution 2D geometry, in order to determine the correct
temperature as a function of cylindrical polar $r$ and $z$
co-ordinates. This is then compared to the temperature from the SPH
disc for each SPH particle. Positive values of the fractional error
indicate that the SPH disc is hotter than the expected value.
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/tdiff_1e8}}
  \caption{Fractional temperature error for discs represented by
    different numbers of SPH particles. Axis units are AU.} 
  \label{fig:n_part_tem_diff}
\end{figure*}
When using $10^5$ particles to represent the disc there is a too much
heating in the disc mid plane and insufficient heating further out of
the mid plane. The effect is reduced as the number of particles is
increased and reaches a very low level with $10^8$ particles.

@@

SEDs are plotted in Fig.~\ref{fig:n_part_sed_013} for a viewing angle
of 12.5 degrees and Fig.~\ref{fig:n_part_sed_077} for a viewing angle
of 77.5 degrees. At the shortest wavelengths the 12.5 degrees
inclination angle SED is dominated by emission from the central
star. This region of the SED is well represented with $10^5$ and
$10^8$ particles but is less well represented with $10^6$ or $10^7$
particles where the central star is obscured. At the longest
wavelengths the disc is optically thin and the SEDs will be well
modelled provided the total disc mass is accurately represented. In
all four cases the total disc mass is well represented and the long
wavelength SED matches the benchmark results. At intermediate
wavelengths the SED is sensitive to the details of the disc
representation and it is clear that accurately modelling the SED from
this disc is challenging, even with $10^8$ particles. 
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_013_1e8}}
  \caption{SEDs at a 12.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_013}
\end{figure*}
\begin{figure*}
  \subfigure[$10^5$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e5}}
  \subfigure[$10^6$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e6}}
  \subfigure[$10^7$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e7}}
  \subfigure[$10^8$ particles]{\includegraphics[scale=0.3]{figures/sed_077_1e8}}
  \caption{SEDs at a 77.5 degree inclination angle for discs represented by different numbers of SPH
    particles. The benchmark result is plotted as a solid line and the
  modelled SED is plotted as a dashed line. } 
  \label{fig:n_part_sed_077}
\end{figure*}

@@

Plots of the disc internal energy are shown in Fig.~\ref{fig:log_u} at
siz different times during the simulation. Figure~\ref{fig:frame1}
shows the initial state before the radiative transfer is switched on
when there is a vertically isothermal equation of state. The second
frame (Fig.~\ref{fig:frame2} is shortly after the raditive transfer
has been switched on. The inner disc has collapsed and there is
heating seen in parts of the disc exposed to radiation from the
central star. In Fig.~\ref{fig:frame3} a scale height enhancement is
present in the disc at a radius of approximately
$3\times10^{15}~\rm{cm}$ which shadows the region immediately behind
causing it to cool. Figure~\ref{fig:frame4} shows several such scale
height enhancements an in each case there is cooling behind the
enhancement. The scale height fluctations are able to affect the
mid-plane temperature due to vertical transport of radiation. In
Fig.~\ref{fig:frame5} the inner region of the disc has settled and the
scale height enhancements are seen only in the outer disc and by the
last frame (Fig.~\ref{fig:frame6}) the disc has reached a steady state
in hydrostatic and radiative equilibrium.
\begin{figure*}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6675}\label{fig:frame1}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6691}\label{fig:frame2}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC6763}\label{fig:frame3}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC7363}\label{fig:frame4}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_DSC9363}\label{fig:frame5}}
  \subfigure[]{\includegraphics[scale=0.3]{figures/u_run2_DSC1024}\label{fig:frame6}}
  \caption{Particle plots showing internal energy in cyclindrical
    polar co-ordinates. Axis units are cm.}
  \label{fig:log_u}
\end{figure*}

The surface density of the disc is shown in Fig.~\ref{fig:sigma} where
the solid line is the vertically isothermal disc shown in
Fig.~\ref{fig:frame1} and the dotted line is the disc at the final
time step (Fig.~\ref{fig:frame6}).  Away from the inner and outer
edges of the disc the surface density is well represented by a power
law profile as a function of radius.
\begin{figure}
  \includegraphics[scale=0.3]{figures/sigma}
  \caption{Surface density as a function of radius.}
  \label{fig:sigma}
\end{figure}
The scale height of the disc was determined by fitting Gaussians to
the vertical density profile in 100 radial bins. The fitted scale
height, as a function of radius, is plotted in
Fig.~\ref{fig:scale_height}.  Within the central 10~AU the density
profile appears flat so it was not possible to fit a meaningful scale
height. The solid line is the isothermal disc and the dotted line is
the disc at the final time step.
\begin{figure}
  \includegraphics[scale=0.3]{figures/scale_height}
  \caption{Disc scale height as a function of radius.}
  \label{fig:scale_height}
\end{figure}
The average temperature within one scale height of the disc mid plane
is plotted in Fig.~\ref{fig:mid_tem}. The solid line is the isothermal
disc and the dotted line is the disc at the final time step.
\begin{figure}
  \includegraphics[scale=0.3]{figures/mid_tem}
  \caption{Average temperature within one scale height of the mid plane as a function of
    radius.}
  \label{fig:mid_tem}
\end{figure}

@@

Although we have found a method for constructing an AMR grid from SPH
particles which is robust, in terms of the total mass on the grid, we
found that it was necessary to add additional grid refinement in the
central region of the disc to allow the inner edge to be well
represented. The inner edge of the disc has a large
effect on the temperature distribution in the disc and on the emergent
SED. Consequently the accuracy of the solution is dependent upon the
accurate representation of a region which constitutes a small fraction
of the disc by volume. As the method combines both SPH and grid
elements it is important that both these components have sufficient
resolution to represent the disc inner edge. This presents a
particular challenge for a SPH code which uses equal mass particles
but is achievable with a large number of particles. Accurately
modelling the SED using this method seems to be considerably more
demanding than accurately representing the temperature structure
within the disc and temperature errors can be reduced to a managable
level with a realistic number of SPH particles in the disc.

The combined SPH-radiative transfer code was able to evolve a
circumstellar disc to a state of radiative and hydrostatic
equilibrium. As the disc settles towards an equilibrium state, vertical
motions are seen which give the appearance of an outward propagating
scale height fluctuation. This is a vertical motion which decays more
rapidly at smaller radii and is seen whether radiative transfer is
included or not. One consequence of these fluctuations is that they
can intecept radiation from the central star which can then be
vertically transported into the mid-plane of the disc. Even thought
the fluctuations are of a small amplitude they exert an influence on
the mid-plane temperature.
